Gravitational wave sensitivity curves

### 2014-07-31 Detectors.tex
Christopher's to-do list

In this section we introduce the detector noise curves used in \ref{app:a}. We begin with a description of the basic operation of detectors. We then discuss ground-based detectors, space-based detectors and PTAs in turn. The latter function somewhat differently than conventional interferometers, so we include a brief introduction to PTA analysis. References for the noise curves used for individual detectors can be found in the relevant subsections, further information about the detectors can be in discovered these.

>>>>

In this section we introduce the detector noise curves used in \ref{app:a}. We begin with a description of the basic operation of detectors. We then discuss ground-based detectors, space-based detectors and PTAs in turn. The latter function somewhat differently than conventional interferometers, so we include a brief introduction to PTA analysis. References for the noise curves used for individual detectors can be found in the relevant subsections, further information about the detectors can be in discovered these. Detectors are frequently upgraded and redesigned, hence, while these curves are believed to be correct at the time of writing, it is best to check for updates from the appropriate science teams before relying on the details given here, although we hope that they shall remain accurate enough for illustrative purposes.

### 2014-07-31 Sources.tex
Christopher's to-do list

Rotating neutron stars are a source of continuous GWs if they possess some degree of axial asymmetry \citep{Abbott2007, Prix2009, Einstein@Home}. The signals are near monochromatic with a frequency twice the rotation frequency of the neutron star, and are a potential source for ground-based detectors. The amplitude of the GWs depends upon the deformation of the neutron star, which itself depends upon the neutron star equation of state. This is currently uncertain \citep{Lattimer2012}. Several known pulsars could be sources for the advanced detectors and upper limits from the initial detectors help to constrain the deformations. The boxes labelled ``\emph{pulsars}'' plotted in \ref{app:a} correspond to the upper limits placed on a GW signal from the Crab pulsar \citep{Aasi2014a}, extrapolated across a frequency range between $(20$--$10^{3})~\mathrm{Hz}$.

>>>>

Rotating neutron stars are a source of continuous GWs if they possess some degree of axial asymmetry \citep{Abbott2007, Prix2009, Einstein@Home}. The signals are near monochromatic with a frequency twice the rotation frequency of the neutron star, and are a potential source for ground-based detectors. The amplitude of the GWs depends upon the deformation of the neutron star. The magnitude of the distortion depends upon the neutron star equation of state and the stiffness of the crust, which are currently uncertain \citep{Chamel2008,Lattimer2012}; there is a maximum deformation that be supported by a given set of material properties for the neutron-star structure. Deformations can also be supported by internal magnetic fields \citep{Haskell2008}; since the magnetic field can be arbitrarily small, there is no lower limit on the size of the deformation. Several known pulsars could be sources for the advanced detectors and upper limits from the initial detectors help to constrain the deformations. The boxes labelled ``\emph{pulsars}'' plotted in \ref{app:a} correspond to the upper limits placed on a GW signal from the Crab pulsar \citep{Aasi2014a}, extrapolated across a frequency range between $(20$--$10^{3})~\mathrm{Hz}$.

### 2014-07-31 Typesofsource.tex
Christopher's to-do list

The form for $h_\mathrm{c}$ can be derived from the Fourier transform in the stationary-phase approximation. Consider a source signal with approximately constant amplitude $h_0$ and central frequency $f'$. In this case,
\begin{eqnarray}
h(t) &=& h_0\exp\left[\rmi \phi(t)\right] \; ; \\
\tilde{h}(f) &=& \int_{-\infty}^{\infty}\rmd t\; h_0\exp \left\{2\pi \rmi \left[\frac{\phi(t)}{2\pi t}-f\right]t\right\} \; .
\end{eqnarray}
Without loss of generality, we can assume an initial phase of zero, such that $\phi(t) \approx 2\pi f't$. The largest contribution to the integral comes from where the argument of the exponential is approximately zero. Expanding the exponent in powers of $t$ about $f = f'$,
\begin{eqnarray} \label{eq:FTofEMRI}
\tilde{h}(f) &\simeq& \int_{-\infty}^{\infty}\rmd t\; h_0 \exp\left[2\pi\rmi \dot{f'} \left(t^{2}+\frac{f'-f}{\dot{f'}}t\right)\right]\nonumber \\
&\simeq& h_0\exp\left[-\pi\rmi\frac{\left(f'-f\right)^{2}}{2\dot{f'}}\right]\int_{-\infty}^{\infty}\rmd t\;\exp\left[2\pi\rmi\dot{f'}\left(t+\frac{f'-f}{2\dot{f'}}\right)^{2}\right] \nonumber \\
&\simeq& \frac{h_0}{\sqrt{2\rmi\dot{f'}}}\exp\left[-\pi\rmi\frac{\left(f'-f\right)^{2}}{2\dot{f'}}\right]\; .
\end{eqnarray}
From (\ref{eq:strain-hc}) and (\ref{eq:FTofEMRI}), the characteristic strain for inspiralling sources is given by \citep{FinnThorne}
\begin{equation}\label{eq:insphc}
h_\mathrm{c}(f) = \sqrt{\frac{2f^{2}}{\dot{f}}}h_0 \;.
\end{equation}
Equation (\ref{eq:hc}) should be considered as the definition of characteristic strain and (\ref{eq:insphc}) a consequence of it for inspirals. Equation (\ref{eq:insphc}) is the relation between $h_\mathrm{c}(f)$ and the instantaneous amplitude $h_0$ for an inspiralling source; for other types of source a new relation satisfying (\ref{eq:hc}) has to be found.

@@

Equation (\ref{eq:differentdescriptions}) gives an expression for the energy density in GWs per logarithmic frequency interval,
\begin{equation}\label{eq:stoch}
fS_{\mathrm{E}}(f)=\frac{\pi c^{2}}{4G}f^{2}h_\mathrm{c}^{2}(f) \; .
\end{equation}
The total energy emitted in the logarithmic frequency interval $\mathrm{d}\left(\log f_{\mathrm{r}}\right)$ by a single binary in the population is $\left[\mathrm{d}E_{\mathrm{GW}}/\mathrm{d}(\log f_{\mathrm{r}})\right]\mathrm{d}(\log f_{\mathrm{r}})$; the energy density may be written as
\begin{equation}\label{eq:Phinney}
fS_{\mathrm{E}}(f) = \int_{0}^{\infty} \rmd z\; \frac{\mathrm{d}\nu}{\mathrm{d}z}\frac{1}{(1+z)}\frac{1}{d_{\mathrm{L}}^{2}(z)}\frac{\mathrm{d}E_{\mathrm{GW}}}{\mathrm{d}\left(\log f_{\mathrm{r}} \right)} \; , \end{equation}
where the factor of $\left( 1+z \right)^{-1}$ accounts for the redshifting of the energy.

For simplicity, consider the background to comprise of binaries in circular orbits, with frequencies $f_\mathrm{GW} =f_{\mathrm{r}}/2$, which are far from their last stable orbit. The energy radiated may then be calculated using the quadrupole approximation \citep{petersmathews1963}. The energy in GWs from a single binary per logarithmic frequency interval is
\begin{equation}\label{eq:Thorne}
\frac{\mathrm{d}E_{\mathrm{GW}}}{\mathrm{d}\left(\log f_{\mathrm{r}} \right)} = \frac{G^{2/3}\pi^{2/3}}{3}{\cal{M}}^{5/3}f_{\mathrm{r}}^{2/3}
\end{equation}
between an minimum and maximum frequency set by the initial and final radius of the binary orbit. Here, we assume that the maximum and minimum frequencies are outside of the range of our detector and hence can be neglected. Using (\ref{eq:stoch}), (\ref{eq:Phinney}) and (\ref{eq:Thorne}), an expression for characteristic strain can now be found \citep{SesanaVecchioColancino}
\begin{equation}\label{eq:bigint}
h_\mathrm{c}^{2}(f) = \frac{4G^{5/3}}{3\pi^{1/3}c^{2}}f^{-4/3}\int_{0}^{\infty}\mathrm{d}z\;\int_{0}^{\infty}\mathrm{d}{\cal{M}}\;\frac{\mathrm{d}^{2}\nu}{\mathrm{d}z\,\mathrm{d}{\mathcal{M}}}\frac{1}{d_{\mathrm{L}}^{2}(z)}\left( \frac{{\mathcal{M}}^{5}}{1+z} \right)^{1/3}\; .
\end{equation}

@@

As discussed in \ref{sec:insp}, the expected relation between $h_\mathrm{c}(f)$ and a typical waveform $\tilde{h}(f)$ is
\begin{equation}\label{eq:simple} 
h_\mathrm{c}(f) = \left|\tilde{h}(f)\right|\sqrt{{N}_{\mathrm{cycles}}} \; , 
\end{equation}
where ${N}_{\mathrm{cycles}}$ is the number of cycles of radiation generated by the source, which is of order unity for bursts.

>>>>

The form for $h_\mathrm{c}$ can be derived from the Fourier transform in the stationary-phase approximation. Consider a source signal with approximately constant (root-mean-square) amplitude $h_0$ and central frequency $f'$. In this case,
\begin{eqnarray}
h(t) &=& \sqrt{2} h_0 \cos\left[\rmi \phi(t)\right] \; ; \\
\tilde{h}(f) &=& \frac{h_0}{\sqrt{2}} \int_{-\infty}^{\infty}\rmd t\; \exp \left\{2\pi \rmi \left[\frac{\phi(t)}{2\pi t}-f\right]t\right\} + \exp \left\{-2\pi \rmi \left[\frac{\phi(t)}{2\pi t}+f\right]t\right\} \; .
\label{eq:fourier-h}
\end{eqnarray}
Without loss of generality, we can assume an initial phase of zero, such that $\phi(0) = 0$. The largest contribution to the integral comes from where the argument of the exponentials is approximately zero. For the first term, this occurs when $f = f'$, then the term in brackets is
\begin{equation}
\left[\frac{\phi(t)}{2\pi t}-f\right]_{f\,=\,f'} = f' + \dot{f'}t + \mathcal{O}\left(t^2\right) - f' = \dot{f'}t + \mathcal{O}\left(t^2\right).
\end{equation}
The higher-order terms cause the expoential to oscillate rapidly such that these terms integrate to zero and may be neglected. Performing a similar expansion about $f = -f'$ for the second term in (\ref{eq:fourier-h}) gives
\begin{eqnarray} \label{eq:FTofEMRI}
\tilde{h}(f) &\simeq& \frac{h_0}{\sqrt{2}} \int_{-\infty}^{\infty}\rmd t\; \exp\left(2\pi\rmi \dot{f'} t^{2}\right) + \exp\left(-2\pi\rmi \dot{f'} t^{2}\right) \nonumber \\
&\simeq& \frac{h_0}{\sqrt{2\dot{f'}}}.
\end{eqnarray}
From (\ref{eq:strain-hc}) and (\ref{eq:FTofEMRI}), the characteristic strain for inspiralling sources is given by \citep{FinnThorne}
\begin{equation}\label{eq:insphc}
h_\mathrm{c}(f) = \sqrt{\frac{2f^{2}}{\dot{f}}}h_0 \;.
\end{equation}
Equation (\ref{eq:hc}) should be considered as the definition of characteristic strain and (\ref{eq:insphc}) a consequence of it for inspirals. Equation (\ref{eq:insphc}) is the relation between $h_\mathrm{c}(f)$ and the instantaneous root-mean-square amplitude $h_0$ for an inspiralling source; for other types of source a new relation satisfying (\ref{eq:hc}) has to be found.

@@

Equation (\ref{eq:differentdescriptions}) gives an expression for the energy density in GWs per logarithmic frequency interval,
\begin{equation}\label{eq:stoch}
fS_{\mathrm{E}}(f)=\frac{\pi c^{2}}{4G}f^{2}\left[h_\mathrm{c}(f)\right]^2 \; .
\end{equation}
The total energy emitted in the logarithmic frequency interval $\mathrm{d}\left(\log f_{\mathrm{r}}\right)$ by a single binary in the population is $\left[\mathrm{d}E_{\mathrm{GW}}/\mathrm{d}(\log f_{\mathrm{r}})\right]\mathrm{d}(\log f_{\mathrm{r}})$; the energy density may be written as
\begin{equation}\label{eq:Phinney}
fS_{\mathrm{E}}(f) = \int_{0}^{\infty} \rmd z\; \frac{\mathrm{d}\nu}{\mathrm{d}z}\frac{1}{(1+z)}\frac{1}{\left[d_{\mathrm{L}}(z)\right]^2}\frac{\mathrm{d}E_{\mathrm{GW}}}{\mathrm{d}\left(\log f_{\mathrm{r}} \right)} \; , \end{equation}
where the factor of $\left( 1+z \right)^{-1}$ accounts for the redshifting of the energy.

For simplicity, consider the background to comprise of binaries in circular orbits, with frequencies $f_\mathrm{GW} =f_{\mathrm{r}}/2$, which are far from their last stable orbit. The energy radiated may then be calculated using the quadrupole approximation \citep{petersmathews1963}. The energy in GWs from a single binary per logarithmic frequency interval is
\begin{equation}\label{eq:Thorne}
\frac{\mathrm{d}E_{\mathrm{GW}}}{\mathrm{d}\left(\log f_{\mathrm{r}} \right)} = \frac{G^{2/3}\pi^{2/3}}{3}{\cal{M}}^{5/3}f_{\mathrm{r}}^{2/3}
\end{equation}
between an minimum and maximum frequency set by the initial and final radius of the binary orbit. Here, we assume that the maximum and minimum frequencies are outside of the range of our detector and hence can be neglected. Using (\ref{eq:stoch}), (\ref{eq:Phinney}) and (\ref{eq:Thorne}), an expression for characteristic strain can now be found \citep{SesanaVecchioColancino}
\begin{equation}\label{eq:bigint}
\left[h_\mathrm{c}(f)\right]^2 = \frac{4G^{5/3}}{3\pi^{1/3}c^{2}}f^{-4/3}\int_{0}^{\infty}\mathrm{d}z\;\int_{0}^{\infty}\mathrm{d}{\cal{M}}\;\frac{\mathrm{d}^{2}\nu}{\mathrm{d}z\,\mathrm{d}{\mathcal{M}}}\frac{1}{\left[d_{\mathrm{L}}(z)\right]^2}\left( \frac{{\mathcal{M}}^{5}}{1+z} \right)^{1/3}\; .
\end{equation}

@@

As discussed in \ref{sec:insp}, the expected relation between $h_\mathrm{c}(f)$ and a typical waveform $\tilde{h}(f)$ is
\begin{equation}\label{eq:simple} 
h_\mathrm{c}(f) = \sqrt{{N}_{\mathrm{cycles}}}\left|\tilde{h}(f)\right| \; , 
\end{equation}
where ${N}_{\mathrm{cycles}}$ is the number of cycles of radiation generated by the source, which is of order unity for bursts.
